---
title: "spatial_hw"
author: "Hannah Houston and Ansley Williamson"
date: '2023-02-28'
output: html_document
---

## Spatial Modeling Assignment

```{r}
library(vegan)
library(nlme)
library(gstat)

data(BCI)
## UTM Coordinates (in metres)
BCI_xy = data.frame(x = rep(seq(625754, 626654, by=100), each=5), 
                    y = rep(seq(1011569,  1011969, by=100), len=50))
```

1) Examine if there is evidence of spatial dependence in a rare and a common
species in the BCI tree dataset

```{r}
# look at species ranks based upon number of occurrences
abundance <- sort(colSums(BCI > 0), dec = T)
abundance
hist(abundance, mainlab= "Species Abundance Distribution")
#common species will be Tetragastris.panamensis (count is at 50) and rare species will be Inga.umbellifera (count is at 12)

#calculate Euclidean distances and plot
rare_dist <- dist(BCI$Inga.umbellifera, method = "euclidean")
common_dist <- dist(BCI$Tetragastris.panamensis, method = "euclidean")
xy_dist <- dist(BCI_xy)

max_dist <- max(xy_dist) / 2

#make a plot to look at spatial signal
plot(xy_dist, common_dist, main = "Spatial Signal between UTM Coordinates and a Common Species")
abline(lm(common_dist ~ xy_dist), lwd=3, col='cyan')
lines(lowess(xy_dist, common_dist), lwd=3, col='red')
abline(v = max_dist, col='red', lwd=3, lty=2)

plot(xy_dist, rare_dist, main = "Spatial Signal between UTM Coordinates and a Rare Species")
abline(lm(rare_dist ~ xy_dist), lwd=3, col='blue')
lines(lowess(xy_dist, rare_dist), lwd=3, col='pink')
abline(v = max_dist, col='pink', lwd=3, lty=2)

# compute correlation
obs_cor_com_rare <- cor(xy_dist, rare_dist)
obs_cor_com_rare #this is 0.0345
#this shows a weak correlation close to 0, suggests a random spatial signal. Still not sold so I will check with permutations

obs_cor_common <- cor(xy_dist, common_dist)
obs_cor_common #this is 0.076
#this shows a weak correlation close to 0, suggests a random spatial signal. Still not sold so I will check with permutations

#carry out mantel test for permutations, does 999 automatically, develops a null distribution of correlation values
rare_mantel <- mantel(xy_dist, rare_dist)
rare_mantel
#not significant, shows a random spatial signal 

common_mantel <- mantel(xy_dist, common_dist)
common_mantel
#not significant, shows a random spatial signal
 
```
2) Build two generalized linear models to predict the abundance of the species
*Drypetes standleyi* using the abundance of other tree species in the study site.
Specifically examine the following species as predictor variables:

```{r}
sp_ids = c("Cordia.lasiocalyx", "Hirtella.triandra",
           "Picramnia.latifolia", "Quassia.amara",
           "Tabernaemontana.arborea", "Trattinnickia.aspera", 
           "Xylopia.macrantha")
```

With both models examine the spatial dependence of the residuals using the 
function `Variogram`. Model the spatial dependence in the residuals using one
of the error structures available. 


* Model 1: only include a single species as a predictor variable
```{r}
#subset the data to include only the species and spatial location we are interested in 
dry_sta <- subset(BCI, select = c("Drypetes.standleyi", "Cordia.lasiocalyx", "Hirtella.triandra",
         "Picramnia.latifolia", "Quassia.amara", "Tabernaemontana.arborea" , "Trattinnickia.aspera", "Xylopia.macrantha"))
bci_sub <- cbind(dry_sta, BCI_xy)

#Making Model 1: only include a single species as a predictor variable 

gls_onesp <- gls(Drypetes.standleyi ~ Cordia.lasiocalyx, data = bci_sub)
summary(gls_onesp)
#p-value is significant, absolute value of correlationis 0.87 (large), does seem to be autocorrelation between Drypetes standleyi and Cordia lasiocalyx

#Use function Variogram for one species 
library(gstat)
vario_onesp <- Variogram(gls_onesp, form= ~ x + y, resType = 'response')
# note the default of the Variogram function is to compute normalized residuals
plot(vario_onesp)

#compare to a plot we make on our own 
res <- residuals(gls_onesp)
res_var <- dist(res)^2 * 0.5
plot(dist(bci_sub[, c('x', 'y')]), res_var)
lines(lowess(dist(bci_sub[, c('x', 'y')]), res_var), col='red', lwd=2)
abline(v = max_dist, col='red', lwd=3, lty=2)
#Variogram and the variogram done by hand have a positive slope, the error of the model is showing positive autocorrelation

#Model the spatial dependence in the residuals using one of the error structures available. 
#try an AR1: most common model of spatial autocorrelation
onesp_exp <- update(gls_onesp, corr=corExp(form=~ x + y))
summary(onesp_exp)

#make a variogram with that, examines fit of raw model residuals to error model
plot(Variogram(onesp_exp, maxDist = max_dist)) #has a super tiny nugget (non 0 y intercept)

#model does not fit error very well, try seeing if residuals should be normally distributed
plot(Variogram(onesp_exp, resType='normalized', maxDist = max_dist))
#that looks like residuals are normally distributed

#address the nugget: exponential 
onesp_exp_nug <- update(onesp_exp, corr=corExp(c(0.5, 0.1), form=~x + y, nugget=T))
plot(Variogram(onesp_exp_nug, maxDist = max_dist)) #not a great fit
plot(Variogram(onesp_exp_nug, resType='n', maxDist = max_dist)) #residuals look normally distributed

#address the nugget: rational quadratic error model
onesp_rat_nug<-update(onesp_exp, corr=corRatio(form=~x + y, nugget=T))
plot(Variogram(onesp_rat_nug, maxDist = max_dist)) #still not a great fit
plot(Variogram(onesp_rat_nug, resType='n', maxDist = max_dist)) #residuals look normally distributed
```

* Model 2: include all of the species as predictor variables
```{r}
#Making Model 2: include all species as predictor variables

gls_prespec <- gls(Drypetes.standleyi ~ Cordia.lasiocalyx + Hirtella.triandra + Picramnia.latifolia + Quassia.amara + Tabernaemontana.arborea + Trattinnickia.aspera + Xylopia.macrantha, data = bci_sub)
summary(gls_prespec)

#use function Variogram for all predictor species 
vario_prespec <- Variogram(gls_prespec, form= ~ x + y, resType = 'response')
# note the default of the Variogram function is to compute normalized residuals
plot(vario_prespec)

#compare to a plot we make on our own 
res <- residuals(gls_prespec)
res_var <- dist(res)^2 * 0.5
plot(dist(bci_sub[, c('x', 'y')]), res_var)
lines(lowess(dist(bci_sub[, c('x', 'y')]), res_var), col='red', lwd=2)
abline(v = max_dist, col='red', lwd=3, lty=2)

#based on the variograms, autocorrelation signal seems to weaken (not as drastic in the variogram and extremely flat slope in the hand made version, indicates a random spatial pattern)

#Model the spatial dependence in the residuals using one of the error structures available. 
#try an AR1: most common model of spatial autocorrelation
prespec_exp <- update(gls_prespec, corr=corExp(form= ~ x + y))
summary(prespec_exp)

#make a variogram with that, examines fit of raw model residuals to error model
plot(Variogram(prespec_exp, maxDist = max_dist))

#model does not fit error very well, try seeing if residuals should be normally distributed
plot(Variogram(prespec_exp, resType='normalized', maxDist = max_dist))
#that looks good

#address the nugget: exponential 
prespec_exp_nug <- update(prespec_exp, corr=corExp(c(0.5, 0.1), form=~x + y, nugget=T))
plot(Variogram(prespec_exp_nug, maxDist = max_dist)) #still not a great fit
plot(Variogram(prespec_exp_nug, resType='n', maxDist = max_dist)) #residuals look normally distributed, maybe a slight trend in them

#address the nugget: rational quadratic error model
prespec_rat_nug<-update(prespec_exp, corr=corRatio(form=~x + y, nugget=T))
plot(Variogram(prespec_rat_nug, maxDist = max_dist)) #eh looks a little better than the other models byt still not a super great fit
plot(Variogram(prespec_rat_nug, resType='n', maxDist = max_dist)) #residuals look normally distributed
```


* Did including the spatial error term have a large impact on the coefficients
of the model? 

In both models 1 (Drypetes.standleyi ~ Cordia.lasiocalyx) and 2 (Drypetes.standleyi ~ specified predictor species), including the spatial error term had a large impact on the coefficients of the model. However, the spatial model with a single predictor generally have stronger predictive power.

* Did including the spatial error terms significantly improve model fit (use 
function `anova` to carry out model comparison)?

```{r}
# use 'anova' to carry out model comparison of gls vs. lm for model 1
anova(gls_onesp, onesp_exp, onesp_exp_nug, onesp_rat_nug, test=F)
#spatial error term did improve model fit (lower AIC for spatial models than GLS)
#Among the spatial models, rational quadratic error model with the nugget and AR1 fit the data equally well (AIC within 2 points) while the exponential model with the nugget did not provide as good of a fit, but still better than the GLS model.


# use 'anova' to carry out model comparison of gls vs. lm for model 2
anova(gls_prespec, prespec_exp, prespec_exp_nug, prespec_rat_nug, test=F)
#including the spatial error term did improve model fit (based off of AIC interpretation, the lower the better the fit)
#Among the spatial models, rational quadratic error model with the nugget and AR1 fit the data equally well (AIC within 2 points) while the exponential model with the nugget did not provide as good of a fit, but still better than the GLS model.

```

* Explain why you did or did not observe a difference in the influence of adding the spatial error term between the two models.

When evaluating Model 1 (Drypetes.standleyi ~ Cordia.lasiocalyx) and Model 2 (Drypetes.standleyi ~ specified predictor species), adding the spatial error term did influence the model by improving model fit. This could be because the models that include the spatial error terms account for spatial and temporal autocorrelation while a standard GLS does not. Biologically, I would expect to see autocorrelation to some degree for tree species due to seed dispersal limitations, so using models like AR1 that account for such spatial trends will provide a better fit for the data.


